---
title: "Using Rseasnap for fun and profit"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using Rseasnap for fun and profit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Quick start

```{r}
# Minimal example included in the package
library(Rseasnap)
example_pipeline_file <- file.path(
  system.file("extdata", package="Rseasnap"),
  "example_pipeline",
  "DE_config.yaml")
pip <- load_de_pipeline(config_file=example_pipeline_file)

# get the DESeq2 object
ds <- get_deseq2(pip)

# get the file tab
file_tab <- get_file_tab(pip)

# plot expression of a gene

# launch the browser
if(interactive()) {
  pipeline_browser(pip)
}
```



# Using Rseasnap with sea-snap pipelines

## Loading a pipeline

Basic usage: assuming that `DE_config.yaml` is the name of the
configuration file of the DE pipeline, load the pipeline with

```{r eval=FALSE}
library(Rseasnap)
pip <- load_de_pipeline(config_file="DE_config.yaml")
```

Using the `pip` object, you now can conveniently extract objects from the
pipeline, create visualisations or launch an interactive shiny app.

## Getting objects from the pipeline

Use the `get_*()` functions to read the RDS files stored in the pipeline
and extract objects:

```{r eval=FALSE}
file_tab <- get_file_tab(pip)
ds2      <- get_deseq2(pip)
cntr     <- get_contrasts(pip)
cntr_ids <- get_contrast_names(pip)
covar    <- get_covariates(pip)
config   <- get_config(pip)
annot    <- get_annot(pip)
tmod     <- get_tmod_res(pip)
tmod_dbs <- get_tmod_dbs(pip)
tmod_map <- get_tmod_mapping(pip)
```

Anything not included in the commands above can be extracted using
`get_object()`. This reads the `file_tab` and extracts object or objects
for the given contrast, step and extension:

```{r eval=FALSE}
obj <- get_object(pip, step="DESeq2", extension="rld.blind.rds")
```

Note that reading of files other than RDS is not supported.

## Storing objects for future use

Some of the objects (notably `tmod_dbs`) are quite large, so if you plan to
repeatedly, say, generate tmod evidence plots or run the pipeline browser,
it makes sens to pre-load them and use them as an optional argument for
other functions, e.g.:

```{r eval=FALSE}
tmod_dbs <- get_tmod_dbs(pip)
pipeline_browser(pip, tmod_dbs=tmod_dbs)
```


# Visualizations

## Transcription profiles

There are two functions for plotting transcription profiles, `plot_gene`
and `plot_gene_generic`, the former specifically for sea-snap pipelines and
the latter for any sort of expression data (`plot_gene` is just a wrapper
around `plot_gene_generic` loading the data first from the pipeline if
necessary).

Both functions need a covariate data frame and a covariate name rather than
a covariate vector, and an expression matrix and gene ID rather than a
numeric vector. Both functions return a ggplot2 object.

If the x covariate is not numeric, then the default plot is
a box plot with a jitter plot overlayed (unless the data is grouped by a
variable, eg. patient ID, in which case it is a simple dot plot with
lines). If the covariate is numeric, then the result is a regular
scatterplot. 

The parameters `symbolBy` and `colorBy` can be used to select covariates
for symbols and colors.

```{r, fig.width=8,fig.height=4}
library(cowplot)
library(ggplot2)
theme_set(theme_bw())
g1 <- plot_gene(pip, "ENSG00000131747", "group.icu")
g2 <- plot_gene(pip, "ENSG00000131747", "fibrinogen",
  colorBy="group", symbolBy="icu")
plot_grid(g1, g2)
```

As these are ggplot2 objects, all the parameters can be modified:

```{r fig.width=4,fig.height=4}
g1$layers[[2]]$aes_params$size <- 1
g1
```


To make these plots faster, load the annotation and expression data first
and provide them to `plot_gene` using the parameters `annot` and `exprs`.

## Evidence plots

Evidence plots are ROC curves for gene set enrichments, produced by the
`evidencePlot` function from the `tmod` package. 

```{r fig.width=6,fig.height=4}
plot_evidence(pip, id="LI.M4.2", dbname="tmod", contrast="COVID19_ID0")
```

## Disco plots

Disco plots visualize a comparison between two contrasts. Genes which
change in the same direction in both contrasts are shown in red, genes with
opposite changes are shown in blue. Color intensity depends on the mapping
between the disco scores and can be modified by the parameters `lower` and
`upper`.

```{r fig.width=10,fig.height=5}
cntr <- get_contrasts(pip)
g1 <- plot_disco(cntr[[1]], cntr[[2]])
g2 <- plot_disco(cntr[[1]], cntr[[2]], lower=-5, upper=5)
plot_grid(g1, g2)
```

The colors follow the disco statistic, defined as 

$$d = \text{LFC}_1\cdot \text{LFC}_2\cdot(-\log_{10}(p_1) - \log_{10}(p_2))$$

The disco score can be calculated with `disco_score()`.

# Functional analysis

## Gene set enrichments with tmod

The sea-snap pipeline uses by default a number of gene set databases,
including the original expression modules from the `tmod` package and gene
sets derived from the MSigDB database. These are stored in the `tmod_dbs`
object (in the pipeline, step `tmod_dbs`, extension `rds`, contrast `all`),
returned by the function `get_tmod_dbs()`. 

Since mapping between the identifiers used by the pipeline and the
databases is a bit tricky, the pipeline exports the mapped identifiers in
the object returned by the `get_tmod_mapping()` function. Using the `tmod`
package for enrichment tests in this context would require a fair amount of
footwork, which is why Rseasnap includes convenience functions for this.

The `test_gsea_tmod()` function requires the pipeline object, a sorted list
of genes (primary identifiers) and the name of the gene set database to
use. The sorted list of genes may be derived from the contrasts results
(i.e., genes sorted by increasing p-value), but there are other options.

For example, one can sort the genes by the decreasing disco score, which
means that the top of the list will consist of most concordant genes
between two contrasts (red dots on a disco plot), while the bottom will be
discordant genes (blue dots).

```{r}
tmod_dbs <- get_tmod_dbs(pip)
disco <- disco_score(cntr[[1]], cntr[[2]])
ord <- order(disco$disco, decreasing=TRUE)
gene_list <- disco$PrimaryID[ ord ]
test_gsea_tmod(pip, gene_list, "tmod")
```


# Pipeline browser

## Usage

Pipeline browser is a simple R shiny app which creates a GUI allowing to
explore the pipeline results. Presently, three modules are available: 

 * a browser for DE analysis results
 * browser for tmod gene set enrichment analysis results, and
 * disco plot viewer

To start the browser in an interactive session, type

```{r eval=FALSE}
pipeline_browser(pip)
```

## Development

Components of the browser are codes as shiny modules, thus enabling
independent usage in other shiny apps. For example, a minimal shiny app
featuring the gene browser (without the profile visualization part!) may look as follows:

```{r eval=FALSE}
library(shiny)
cntr  <- get_contrasts(pip)
annot <- get_annot(pip)

ui <- fluidPage(
  geneBrowserTableUI("geneT", names(cntr))
  )

server <- function(input, output, session) {
  geneBrowserTableServer("geneT", cntr, annot)
}

shinyApp(ui, server)
```




